Geometrization of characters of tori over non-archimedean local fields

### 2013-08-15 shanghai.tex
Part 2 is finished, pending results on base change from Part 1.

Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
\[
\Hh^1(\Weil{K}, \hat{T}) \cong \Hom(T(K), \EEx).
\]
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.

@@

  We now fix a $d$, $P$ and $s$, write $\zeta$ for $\zeta_d$ and
  choose a polynomial
  $$h(T) = a_rT^r + a_{r-1}T^{r-1} + \cdots + a_0$$
  so that $P^s = (p^s, h(\zeta_d))$.
  % Setting $\psi = 1$ will satisfy equation \ref{eq:psi-condition}
  % when $d = 1$.
  An an abelian group, the quotient $\ZZ[\zeta] / P^s$ is generated by
  $\zeta^i$ for $0 \le i < \phi(d)$; we define
  \begin{equation} \label{eq:psi-def}
    \psi(\zeta^i) = \alpha \prod_{j=0}^{i-1} \tilde{f}(\zeta^j)
  \end{equation}
  for an $\alpha \in \EEx$ to be chosen later.  Since there are
  relations among the $\zeta^i$ we first need to check that $\psi$ is
  well-defined.

@@

  We now check equation \ref{eq:psi-condition} for
  $x = \zeta^{\phi(d)-1}$.  Write
  $\Phi_d(T) = T^{\phi(d)} - b_{\phi(d)-1}T^{\phi(d)-1} - \cdots - b_0$
  and set
  $$B = \sum_{i=0}^{\phi(d)-1} -\zeta^i + b_i \sum_{j=0}^{i-1} \zeta^j.$$
  We have

@@

  \begin{enumerate}
  \item $d$ is a power of $p$, in which case $\Phi_d(1) = p$.  Thus
    $(\zeta-1)B = p = u(\zeta-1)^{\phi(d)}$ for a unit $u \in \ZZ[\zeta]$
    and $B - u(\zeta-1)^{\phi(d)-1}$ is fixed by $\zeta$,
    so that $\tilde{f}(B) = \tilde{f}(u(\zeta - 1)^{\phi(d)-1})$.  The
    multiplicative order of $\tilde{f}(u(\zeta-1)^{\phi(d)-1})$ is at most
    $\lceil \frac{s - 1 - (\phi(d)-1)}{\phi(d)}\rceil = \lceil \frac{s}{\phi(d)} \rceil - 1$
    while the additive order of $1 \in \ZZ[\zeta]/P^s$ is $\lceil \frac{s}{\phi(d)} \rceil$.
    Thus we may choose $\alpha = \psi(1)$ with $\alpha^p = \tilde{f}(B)$.
  \item $d$ is not a power of $p$ and $\omega$ is the inverse of
    $\zeta-1$ modulo $P^s$.  Then $\tilde{f}(B) = \tilde{f}(\omega\Phi_d(1))$.
  \end{enumerate}
  In either case, $\alpha^{-\Phi_d(1)}\tilde{f}(B) = 1$ and thus
  $\psi$ satisfies equation \ref{eq:psi-condition} for
  $x = \zeta^{\phi(d)-1}$.  The definition of $\psi$ in \ref{eq:psi-def}
  immediately implies that $\psi$ satisfies equation
  \ref{eq:psi-condition} for lesser powers of $\zeta$, and thus for
  all $x$ by linearity.  The existence of $\psi$ shows that $\rho$ is injective.
\end{proof}

@@

\begin{corollary}\label{cor:application}
\begin{enumerate}
\item 
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T}:  \QCiso{\GN{T}} \to \Hom_{\text{qc}}(T(K),\EEx)
  \]
is an isomomorphism. 
\item The isomorphism $\trFrob{T}$ is functorial in the following sense: $T \mapsto \trFrob{T}$ is a natural transformation
from the additive functor
\[T \mapsto \QCiso{\GN{T}}\]
to the additive functor
\[
T \mapsto \Hom_{\text{qc}}(T(K),\EEx).
\]
\item
For every finite Galois extension $L/K$ and for every algebraic torus $T$ over $K$, the following diagram commutes:
\[
\begin{tikzcd}
\QCiso{\Gr(\underline{\Res_{L/K}(T_L)})} \arrow[two heads]{r} \dar & \QCiso{\Gr({T})} \dar\\
\Hom_{\text{qc}}(T(L), \EEx) \arrow[two heads]{r} & \Hom_{\text{qc}}(T(K), \EEx)
\end{tikzcd}
\]
\end{enumerate}
\end{corollary}

@@

\David{We should add a discussion of why we care about the transfer principle.}

@@

We recall the notion of $N$-congruent tori from  \cite{CY}*{\S 2}.  Suppose $T$ and $T'$
are tori over non-archimedian local fields $K$ and $K'$, splitting over $L$ and $L'$ respectively.
Then $T$ and $T'$ are said to be \emph{$N$-congruent} if there are isomorphisms
 \begin{align*}
  \alpha : \OL/\pi_K^N \OL &\to \OO{L'}/\pi_{K'}^N \OO{L'} \\
  \beta : \Gal(L/K) &\to \Gal(L'/K') \\
  \phi : X^*(T) &\to X^*(T')
 \end{align*}
 satisfying the following conditions:
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\pi_K^N \OK \to \OO{K'}/\pi_{K'}^N \OO{K'}$,
  \item $\alpha$ is $\Gal(L/K)$-equivariant relative to $\beta$: $\alpha(\gamma x) = \beta(\gamma) \alpha(x)$
  for $\gamma \in \Gal(L/K)$ and $x \in \OL/\pi_K^N \OL$,
  \item $\phi$ is $\Gal(L/K)$-equivariant relative to $\beta$.
 \end{enumerate}
\[
  \Hom_{<N}(T(K), \EEx) \to \Hom_{<N}(T'(K'),\EEx)
 \]

\begin{corollary}
 Suppose that $T$ and $T'$ are $N$-congruent for $N > 3h$.  Then there is a canonical equivalence of categories
 \[
  \QC(\GN{T}_{N-3h}) \to \QC(\GN{T}'_{N-3h})
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing the isomorphism
 \[
  \Hom(T(K) / T(K)_{N-3h}, \EEx) \to \Hom(T'(K') / T'(K')_{N-3h}, \EEx)
 \]
\end{corollary}

\begin{proof}
One of the main results of \cite{CY} gives an isomorphism of group schemes between
$\TT_d$ and $\TT'_d$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{CY}*{\S 8.1}
They then prove the following theorem \cite{CY}*{Thm. 8.5}
 Suppose that $T$ and $T'$ are $N$-congruent for $N > 3h$.  Then there is a canonical isomorphism
 \[
  \TT_{N-3h} \to \TT'_{N-3h}
 \]
 determined by $\alpha, \beta$ and $\phi$.

As an immediate consequence of this theorem and Theorem \ref{thm:application} we have the following corollary:
\end{proof}

@@

As indicated in the Introduction, we are tantalized by the idea that quasicharacter sheaves for $T$ may play a key role in a geometric, categorial version of the local Langlands correspondence for algebraic tori over non-archimedean local fields. We anticipate that, as with the local Langlands correspondence for algebraic tori, it will be possible to reduce the essential points of the argument to the case $T= \Gm{K}$, provided that the category of quasicharacter sheaves behaves properly with regard to Weil restriction; see \cite{Yu:tori}*{Thm 7.5 (2)}. The next proposition shows that the category of quasicharacter sheaves does indeed behave properly with regard to Weil restriction.

@@

\begin{proof}
\Clifton{Under construction, pending the proofs of the results in the base change section}
There is a canonical isomorphism
\[
\Res_{k'/k} \Gr(\TL) \cong \Gr(\underline{\Res_{L/K}(T_L)}).
\]
\end{proof}

This paper has shown how quasicharacter sheaves on the Greenberg transform of the Neron model of $T$ determine admissible characters of $T(K)$, and vice versa, functorially. When coupled with the reciprocity map $\text{rec}_T $ for $T$, which enjoys the same functorial properties, it follows that quasicharacter sheaves for $T$ determine Langlands parameters for $T$, and vice versa, functorially.
\[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr}  & \\
\Hom_{<\infty}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}) \\
\end{tikzcd}
\]
We suspect that it is possible to extract Langlands parameters from quasicharacter sheaves directly; results in this paper reduce the problem to the case $T=\Gm{K}$. In that case, our suspicions are all but confirmed by the class field theory of Serre-Hasewinkel \cite{Serre:isogenies} as revisited in \cite{Suzuki&Yoshida:refinement}. What remains is a careful comparison of quasicharacter sheaves for $\Gm{K}$ and certain local systems on the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{Suzuki&Yoshida:refinement}.

@@

\bib{SGA1}{collection}{
   title={Rev\^etements \'etales et groupe fondamental (SGA 1)},
%   language={French},
   author={Grothendieck, A.},
   series={Documents Math\'ematiques (Paris), 3},
   note={S\'eminaire de g\'eom\'etrie alg\'ebrique du Bois Marie 1960--61.
   Un s\'eminaire dirig\'e par A. Grothendieck
Augment\'e de deux expos\'es de Mme M. Raynaud
\'Edition recompos\'e et annot\'e du volume 224 des Lecture Notes in Mathematics 224 publi\'e en 1971 par Springer-Verlag},
   publisher={Soci\'et\'e Math\'ematique de France},
   place={Paris},
   date={2003},
   pages={xviii+327},
%  isbn={2-85629-141-4},
%   review={\MR{2017446 (2004g:14017)}},
}

>>>>

Finally, we are tantalized by the possibility of providing an alternate perspective on local class field theory
through quasicharacter sheaves.  The local Langlands correspondence for tori provides an isomorphism
\[
\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \Hom(T(K), \EEx).
\]
A geometric description of the resulting isomorphism $\Hh^1(\Weil{K}, \hat{T}_\ell) \cong \QCiso{\GN{T}}$
would yield a different description of the Artin reciprocity map.

@@

  We now fix a $d$, $P$ and $s$, write $\zeta$ for $\zeta_d$ and
  choose a polynomial
  $$h(T) = a_rT^r + a_{r-1}T^{r-1} + \cdots + a_0$$
  so that $P^s = (p^s, h(\zeta_d))$.
  % Setting $\psi = 1$ will satisfy Equation \ref{eq:psi-condition}
  % when $d = 1$.
  An an abelian group, the quotient $\ZZ[\zeta] / P^s$ is generated by
  $\zeta^i$ for $0 \le i < \phi(d)$; we define
  \begin{equation} \label{eq:psi-def}
    \psi(\zeta^i) = \alpha \prod_{j=0}^{i-1} \tilde{f}(\zeta^j)
  \end{equation}
  for an $\alpha \in \EEx$ to be chosen later.  Since there are
  relations among the $\zeta^i$ we first need to check that $\psi$ is
  well-defined.

@@

  We now check Equation~\eqref{eq:psi-condition} for
  $x = \zeta^{\phi(d)-1}$.  Write
  $\Phi_d(T) = T^{\phi(d)} - b_{\phi(d)-1}T^{\phi(d)-1} - \cdots - b_0$
  and set
  $$B = \sum_{i=0}^{\phi(d)-1} -\zeta^i + b_i \sum_{j=0}^{i-1} \zeta^j.$$
  We have

@@

  \begin{enumerate}
  \item $d$ is a power of $p$, in which case $\Phi_d(1) = p$.  Thus
    $(\zeta-1)B = p = u(\zeta-1)^{\phi(d)}$ for a unit $u \in \ZZ[\zeta]$
    and $B - u(\zeta-1)^{\phi(d)-1}$ is fixed by $\zeta$,
    so that $\tilde{f}(B) = \tilde{f}(u(\zeta - 1)^{\phi(d)-1})$.  The
    multiplicative order of $\tilde{f}(u(\zeta-1)^{\phi(d)-1})$ is at most
    $\lceil \frac{s - 1 - (\phi(d)-1)}{\phi(d)}\rceil = \lceil \frac{s}{\phi(d)} \rceil - 1$
    while the additive order of $1 \in \ZZ[\zeta]/P^s$ is $\lceil \frac{s}{\phi(d)} \rceil$.
    Thus we may choose $\alpha = \psi(1)$ with $\alpha^p = \tilde{f}(B)$.
  \item $d$ is not a power of $p$ and $\omega$ is the inverse of
    $\zeta-1$ modulo $P^s$.  Then $\tilde{f}(B) = \tilde{f}(\omega\Phi_d(1))$.
  \end{enumerate}
  In either case, $\alpha^{-\Phi_d(1)}\tilde{f}(B) = 1$ and thus
  $\psi$ satisfies Equation~\eqref{eq:psi-condition} for
  $x = \zeta^{\phi(d)-1}$.  The definition of $\psi$ in \ref{eq:psi-def}
  immediately implies that $\psi$ satisfies equation
  \ref{eq:psi-condition} for lesser powers of $\zeta$, and thus for
  all $x$ by linearity.  The existence of $\psi$ shows that $\rho$ is injective.
\end{proof}

@@

\begin{corollary}\label{cor:application}
\begin{enumerate}
\item 
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T}:  \QCiso{\GN{T}} \to \Hom_{\text{qc}}(T(K),\EEx)
  \]
is an isomomorphism. 
\item The isomorphism $\trFrob{T}$ is functorial in the following sense: $T \mapsto \trFrob{T}$ is a natural transformation
from the additive functor
\[T \mapsto \QCiso{\GN{T}}\]
to the additive functor
\[
T \mapsto \Hom_{\text{qc}}(T(K),\EEx).
\]
\end{enumerate}
\end{corollary}

@@

@@

We recall the notion of $N$-congruent tori from \cite{CY}*{\S 2}.  
Suppose $T$ and $T'$ are tori over non-archimedian local fields $K$ and $K'$, 
splitting over $L$ and $L'$ respectively.
Then $T$ and $T'$ are said to be \emph{$N$-congruent} if there are isomorphisms
 \begin{align*}
  \alpha : \OL/\pi_K^N \OL &\to \OO{L'}/\pi_{K'}^N \OO{L'} \\
  \beta : \Gal(L/K) &\to \Gal(L'/K') \\
  \phi : X^*(T) &\to X^*(T')
 \end{align*}
 satisfying the following conditions:
 \begin{enumerate}
  \item $\alpha$ induces an isomorphism $\OK/\pi_K^N \OK \to \OO{K'}/\pi_{K'}^N \OO{K'}$,
  \item $\alpha$ is $\Gal(L/K)$-equivariant relative to $\beta$: $\alpha(\gamma x) = \beta(\gamma) \alpha(x)$
  for $\gamma \in \Gal(L/K)$ and $x \in \OL/\pi_K^N \OL$,
  \item $\phi$ is $\Gal(L/K)$-equivariant relative to $\beta$.
 \end{enumerate}
It follows from this definition that if $T$ and $T'$ are $N$-congruent then $\alpha$, $\beta$ and $\phi$ determine an isomorphism
\begin{equation}\label{transfer}
  \Hom_{<N}(T(K), \EEx) \iso \Hom_{<N}(T'(K'),\EEx).
\end{equation}
Note that if $T$ and $T'$ are $N$-congruent, then they are $N'$ congruent for every $N' \leq N$.
In this section we ask if the isomorphism \eqref{transfer} comes from an equivalence of categories of quasicharacter sheaves on $\GN{T}_d$ and those on $\GN{T'}_d$.

One of the main results of \cite{CY} gives an isomorphism of group schemes between
$\TT_d$ and $\TT'_d$ assuming that $T$ and $T'$ are sufficiently congruent.
They define a quantity $h$ as the smallest integer so that $\pi^h$ lies in the
Jacobian ideal associated to a natural embedding of $T$ into an induced torus \cite{CY}*{\S 8.1}; let $h'$ be this quantity for $T'$.
Then it follows from \cite{CY}*{Thm. 8.5} that if $T$ and $T'$ are $N$-congruent and $h=h'$ and $N > 3h$, then there is a canonical isomorphism
 \[
  \TT_{N-3h} \to \TT'_{N-3h}
 \]
 determined by $\alpha, \beta$ and $\phi$.
As an immediate consequence of this theorem and Theorem \ref{thm:application} we have the following

\begin{proposition}
 With notation above, suppose that $T$ and $T'$ are $N$-congruent and $h=h'$ and $N > 3h$.  Set $d = N-3h$.
 Then there is a canonical equivalence of categories
 \[
  \QC(\GN{T}'_{d}) \to \QC(\GN{T}_{d})
 \]
 determined by $\alpha, \beta$ and $\phi$ inducing the isomorphism
 \[
  \Hom_{<d}(T'(K'), \EEx) \to \Hom_{<d}(T(K), \EEx).
 \]
\end{proposition}

@@

As indicated in the Introduction, we are tantalized 
by the idea that quasicharacter sheaves for $T$ 
may play a key role in a geometric, categorial version 
of the local Langlands correspondence for algebraic tori 
over non-archimedean local fields. 
As with the local Langlands correspondence for algebraic tori, 
it will be possible to reduce the essential points of the argument to the case $T= \Gm{K}$, provided that the category of quasicharacter sheaves behaves properly with regard to Weil restriction; see \cite{Yu:tori}*{Thm 7.5 (2)}. 
The next proposition shows that the category of quasicharacter sheaves for 
algebraic tori does indeed behave properly with regard to Weil restriction.

@@

\begin{proof}
\Clifton{The proof will require the results in the base change section.}
There is a canonical isomorphism
\[
\Res_{k'/k} \Gr(\TL) \cong \Gr(\underline{\Res_{L/K}(T_L)}).
\]
\end{proof}

This paper has shown how quasicharacter sheaves 
on the Greenberg transform of the Neron model of $T$ 
determine admissible characters of $T(K)$, and vice versa, functorially. When coupled with the reciprocity map $\text{rec}_T $ for $T$, 
which enjoys the same functorial properties, 
it follows that quasicharacter sheaves for $T$ 
determine Langlands parameters for $T$, and vice versa, functorially.
\[
\begin{tikzcd}
\ & \arrow[swap]{dl}{\trFrob{G}} \QCiso{\GN{T}} \arrow[dashed]{dr}  & \\
\Hom_{<\infty}(T(K),\EEx) \arrow{rr}{\text{rec}_T} && H^1(K,\hat{T}_\ell) \\
\end{tikzcd}
\]
We suspect that it is possible to extract Langlands parameters 
from quasicharacter sheaves directly and that the reciprocity map 
can then be be viewed as pairing characters and parameters coming
 from the same quasicharacter sheaf. 
 Results in this paper reduce the problem to the case $T=\Gm{K}$. 
 In that case, our suspicions are all but confirmed 
 by the class field theory of Serre-Hasewinkel as revisited in \cite{Suzuki&Yoshida:refinement}. 
 What remains to be done is a careful comparison of 
 quasicharacter sheaves for $\Gm{K}$ and certain local systems on
  the fpqc-site of the pro-quasi-algebraic varieties appearing in \cite{Suzuki&Yoshida:refinement}.

@@

\bib{SGA1}{collection}{
   title={Rev\^etements \'etales et groupe fondamental (SGA 1)},
%   language={French},
   author={Grothendieck, A.},
   series={Documents Math\'ematiques (Paris), 3},
   note={S\'eminaire de g\'eom\'etrie alg\'ebrique du Bois Marie 1960--61.
   Un s\'eminaire dirig\'e par A. Grothendieck
Augment\'e de deux expos\'es de Mme M. Raynaud
\'Edition recompos\'e et annot\'e du volume 224 des Lecture Notes in Mathematics publi\'e en 1971 par Springer-Verlag},
   publisher={Soci\'et\'e Math\'ematique de France},
   place={Paris},
   date={2003},
   pages={xviii+327},
%  isbn={2-85629-141-4},
%   review={\MR{2017446 (2004g:14017)}},
}
